Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2022-03-14 writing/investigating-GPP.tex
started merging appendix C and D

\section{Control stimuli in the projection blocks of Exps.~1-3}\label{a-control}

The control stimuli in Exps.~1-3 were main clauses. In Exps.~1n and 1m, we used the sentences in

@@

\begin{exe}
\exi{(1)}  Sentences for control stimuli in Exps.~1
\begin{xlist}
\ex These muffins have blueberries in them.  (NRRC: , which are really delicious, )
\ex This pizza has mushrooms on it. (NRRC: , which I just made from scratch, )
\ex Jack was playing outside with the kids. (NRRC: , who is my long-time neighbor, )
\ex Ann dances ballet. (NRRC: , who is a local performer, )
\ex John's kids were in the garage. (NRRC: , who are very well-behaved, )
\ex Samantha has a new hat. (NRRC: , who is really into fashion, )
\end{xlist}
\end{exe}
We expected participants to give low responses in the `certain that'  diagnostic  in Exp.~1q (indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question) and high responses in the `certain that'  diagnostic in Exps.~1n, 1m, and 1c (indicating that the speaker is certain of the main clause content). These expectations were borne out, as shown in the first three rows of Table \ref{t-proj-controls}.

%We used positive declarative control sentences in Exps.~1n, 1m, and 1c so that we could have clear expectations about participants' responses. Thus, we expected participants to give low responses in the `certain that' and `asking whether' diagnostic in Exp.~1q (indicating that the main clause content does not project and is at-issue), and we expected participants to give low responses in the `certain that' diagnostic and high responses in the `sure that' diagnostic in Exps.~1n, 1m, and 1c (indicating, again, that the main clause content does not project and is at-issue).

In Exps.~2 and 3, the NRRCs were included in all control stimuli: the carrier sentences were the polar question variants of the sentences in (1) in Exps.~2q and 3q, and the sentences in (1) for the remaining experiments. The NRRCs were included in Exps.~2 and 3 to make the use of the assent diagnostics in the at-issueness blocks more natural: like the target stimuli, the control stimuli consist of two clauses, one of which the relevant speaker assents with.

We had the same expectations as for the control stimuli in Exps.~1: low responses in the `certain that' diagnostic in Exps.~2q and 3q (indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question), and high responses in the `certain that' diagnostic in the remaining experiments (indicating, again, that the speaker is certain of the main clause content). These expectations were borne out, as shown in the lower nine rows of Table \ref{t-proj-controls}.

\begin{table}[h!]
\centering
\begin{tabular}{r r}
Experiment & mean certainty rating \\ 
\hline
1q & .14 \\
2q & .18 \\
3q & .17 \\
\hline
1n & .95 \\
1m & .96 \\
1c & .94 \\
\hline
2n & .96 \\
2m & .96 \\
2c & .96 \\
\hline
3n & .94 \\
3m & .93 \\
3c & .93 \\
\hline
\end{tabular}
\caption{Mean certainty ratings for control stimuli, for self-declared American English participants}\label{t-proj-controls}
\end{table}

\section{Ratings for main clause contents in at-issueness blocks in Exps.~1-3}\label{a-control-ai}

In the at-issueness blocks of Exps.~1-3, we collected at-issueness ratings for the main clause contents of the control stimuli of the projection blocks described in Supplement \ref{a-control}. Originally, the intent was to exclude participants' data on the basis of their at-issueness ratings, in parallel to their certainty ratings. We expected the main clause content to be at-issue across the diagnostics, which means that we expected low responses across all at-issueness diagnostics (recall that participants' responses were coded so that the higher a participants' response, the more not-at-issue the content was hypothesized to be, to investigate whether there is a positive correlation between projection and not-at-issueness). As shown in Table \ref{t-ai-controls}, these expectations were not always borne out, which is why we decided to not use participants' ratings of the main clause contents in the at-issueness blocks as an exclusion criterion.

\begin{table}[h!]
\centering
\begin{tabular}{r l}
Experiment & mean rating \\ 
\hline
1q & .05 \\
2q & .07 \\
3q & .28 \\
\hline
1n & .04 \\
1m & .03 \\
1c & .08 \\
\hline
2n & .22 (16 participants excluded by specified exclusion criterion) \\
2m & .25 \\
2c & .22 \\
\hline
3n & .44  (nobody excluded by specified exclusion criterion) \\
3m & .50 \\
3c & .53 \\
\hline
\end{tabular}
\caption{Mean ratings on at-issueness diagnostics for main clause contents in Exps.~1-3, for self-declared American English participants}\label{t-ai-controls}
\end{table}

>>>>

\section{Control stimuli in Exps.~1-3}\label{a-control}

The control stimuli in Exps.~1-3 were main clauses. For the question embedding (Exps.~1q, 2q and 3q), we used the polar questions in (2). The NRRCs in parentheses are only used in Exps.~2q and 3q to make the use of the assent diagnostics in the at-issueness blocks more natural: like the target stimuli, the control stimuli consist of two clauses, one of which the relevant speaker assents with.

\begin{exe}
\exi{(1)} Sentences for control stimuli in question embedding (Exps.~1q, 2q and 3q)
\begin{xlist}
\ex Do these muffins (, which are really delicious,) have blueberries in them?
\ex Does this pizza (, which I just made from scratch,) have mushrooms on it? 
\ex Was Jack (, who is my long-time neighbor,) playing outside with the kids? 
\ex Does Ann (, who is a local performer,) dance ballet?
\ex Were John's kids (, who are very well-behaved,) in the garage?
\ex Does Samantha (, who is really into fashion,) have a new hat?
\end{xlist}
\end{exe}

In the negation, modal and conditional embeddings, we used the positive declarative variants of (2), given in (1). Exps.~1n and 1m, we used the sentences in (1) without the NRRCs. We used the sentences in (1) with the NRRCs for Exp.~1c and for Exps.~2 and 3.

@@

\begin{exe}
\exi{(2)}  Sentences for control stimuli in negation, modal and conditional embeddings
\begin{xlist}
\ex These muffins (, which are really delicious,) have blueberries in them.
\ex This pizza (, which I just made from scratch,) has mushrooms on it. 
\ex Jack (, who is my long-time neighbor,) was playing outside with the kids. 
\ex Ann (, who is a local performer,) dances ballet.
\ex John's kids (, who are very well-behaved,) were in the garage.
\ex Samantha (, who is really into fashion,) has a new hat.
\end{xlist}
\end{exe}

\paragraph{Expectations of answers to certainty and at-issueness diagnostics} For the control stimuli in (1) used in Exps.~1q, 2q and 3q, which are polar questions, we expected participants to give low responses in the `certain that' diagnostic, indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question. For the control stimuli in (2), which are positive declarative sentences, we expected high responses in the `certain that' diagnostic for the control stimuli, indicating that the speaker is certain of the main clause content. These expectations were borne out, as shown in the third column of Table~\ref{t-controls}.

%In Exps.~2 and 3, the NRRCs were included in all control stimuli: the carrier sentences were the polar question variants of the sentences in (1) in Exps.~2q and 3q, and the sentences in (1) for the remaining experiments. 
%We had the same expectations as for the control stimuli in Exps.~1: low responses in the `certain that' diagnostic in Exps.~2q and 3q (indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question), and high responses in the `certain that' diagnostic in the remaining experiments (indicating, again, that the speaker is certain of the main clause content). These expectations were borne out, as shown in the lower nine rows of Table \ref{t-proj-controls}.

\begin{table}[h!]
\centering
\begin{tabular}{r r r r l }
Experiment & Control stimuli & Certainty &  \multicolumn{2}{l}{At-issueness} \\ & & Mean rating & Mean rating & Measure \\ 
\hline
1q & (1) & .14 & .05  & asking whether $c$ \\
2q & (1) with NRRC & .18 & .07 & {\em yes}, $c$\\
3q & (1) with NRRC & .17 &  .28 & {\em yes}, but $\neg c'$\\
\hline
1n & (2) &  .95 & .04 & sure that $c$\\
1m & (2) & .96 & .03 & sure that $c$\\
1c & (2) with NRRC & .94  & .08 & sure that $c$\\
\hline
2n & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
2m & (2) with NRRC& .96 & .25 & {\em yes, that's true}, $c$\\
2c & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
\hline
3n & (2) with NRRC & .94 & .44 & {\em yes, that's true}, but $\neg c'$\\
3m & (2) with NRRC & .93 & .50 & {\em yes, that's true}, but $\neg c'$\\
3c & (2) with NRRC & .93 & .53 & {\em yes, that's true}, but $\neg c'$\\
\hline
\end{tabular}
\caption{Mean certainty and at-issueness ratings for control stimuli, for self-declared American English participants}\label{t-controls}
\end{table}

For the at-issueness diagnostics, we expected the main clause content to be at-issue across the diagnostics. We thus expected low responses across all at-issueness diagnostics, given that participants' responses were coded so that the higher a participants' response, the more not-at-issue the content was hypothesized to be (since we investigate whether there is a positive correlation between projection and not-at-issueness). However, as shown in the fourth column of Table~\ref{t-proj-controls}, the at-issueness ratings are not always low. In particular, xx

Exp.~3q\\
A: Do these muffins, which are really delicious, have blueberries in them?\\
B: Yes, but they aren't really delicious.

Exp.~2n, 2m, 2c\\
A: These muffins, which are really delicious, have blueberries in them.\\
B: Yes, that's true, they have blueberries in them.

Exp.~3n, 3m, 3c\\
A: These muffins, which are really delicious, have blueberries in them.\\
B: Yes, that's true, but they aren't really delicious.

%Originally, the intent was to exclude participants' data on the basis of their at-issueness ratings, in parallel to their certainty ratings. We expected the main clause content to be at-issue across the diagnostics, which means that we expected low responses across all at-issueness diagnostics (recall that participants' responses were coded so that the higher a participants' response, the more not-at-issue the content was hypothesized to be, to investigate whether there is a positive correlation between projection and not-at-issueness). As shown in Table \ref{t-ai-controls}, these expectations were not always borne out, which is why we decided to not use participants' ratings of the main clause contents in the at-issueness blocks as an exclusion criterion.

%\begin{table}[h!]
%\centering
%\begin{tabular}{r l}
%Experiment & mean rating \\ 
%\hline
%1q & .05 \\
%2q & .07 \\
%3q & .28 \\
%\hline
%1n & .04 \\
%1m & .03 \\
%1c & .08 \\
%\hline
%2n & .22 (16 participants excluded by specified exclusion criterion) \\
%2m & .25 \\
%2c & .22 \\
%\hline
%3n & .44  (nobody excluded by specified exclusion criterion) \\
%3m & .50 \\
%3c & .53 \\
%\hline
%\end{tabular}
%\caption{Mean ratings on at-issueness diagnostics for main clause contents in %Exps.~1-3, for self-declared American English participants}\label{t-ai-controls}
%\end{table}
